---
title: "Portfolio: Data Analysis in R"
output: rmarkdown::github_document
---

```{r include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggthemes)
library(dslabs)
```

```{r echo=FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```

This is a compiled showcase of data analysis work I have written in R over the years. I love doing with with `tidyverse` and `ggplot2`, and rely heavy on its environment.

Here are some example dataset analyses and my compilation of frequent data analysis problems plus their solutions.

# 1. Example Data Analysis

## Gapminder Dataset

About Gapminder: "Gapminder identifies systematic misconceptions about important global trends and proportions and uses reliable data to develop easy to understand teaching materials to rid people of their misconceptions."

What was the life expectancy in Germany for the last 60 years?

```{r}
gapminder %>% 
  filter(country == "Germany") %>% 
  ggplot() + 
  geom_line(aes(year,life_expectancy)) +
  scale_y_continuous(labels = scales::comma) + 
  xlab("Year") + ylab("Life Expectancy") + 
  ggtitle("Life Expectancy in Germany from 1960 to 2015")
```

What was the life expectancy in several different countries for the last 60 years?

```{r}
gapminder %>% 
  filter(country %in% c("Germany","China", "Nigeria", "Canada", "Thailand", "Russia")) %>% 
  ggplot() + 
  geom_line(aes(year,life_expectancy, group = country, color = country)) +
  scale_y_continuous(labels = scales::comma) + 
  xlab("Year") + ylab("Life Expectancy") + 
  ggtitle("Life Expectancy in Several Different Countries from 1960 to 2015")
  ggtitle("Life Expectancy in Germany from 1960 to 2015")
```

What are the differences in infant mortality rates by continent?

```{r}
gapminder %>% 
  filter(year == 2014) %>% 
  ggplot() + geom_boxplot(aes(continent, infant_mortality, color = continent), show.legend = F) + 
  ggtitle("Infant Mortality Rates by Continents") + 
  xlab("Continents") + 
  ylab("Infant Mortality Rates")
  
```

What regions in gapminder are there?

```{r}
levels(gapminder$region)
```

Population bar graph of Europe by countries.

```{r}
gapminder %>% 
  filter(year == 2015 & continent == "Europe") %>% 
  ggplot(aes(y=reorder(country, population),x = population)) + 
  geom_bar(stat = "identity") + 
  scale_x_continuous(labels = scales::comma) + 
  ggtitle("Population of European Countries") + 
  xlab("Population") + 
  ylab("Country")
```

Pie chart of Europe's population

```{r}
gapminder %>% 
  filter(year == 2015 & continent == "Europe") %>% 
  ggplot(aes(y=reorder(country, population),x = population)) + 
  geom_bar(stat = "identity") + 
  scale_x_continuous(labels = scales::comma) + 
  coord_polar("y", start=0) +
  ggtitle("Population of European Countries") + 
  xlab("Population") + 
  ylab("Country")
```

What are the countries with the biggest life expectancy in Europe?

```{r}
gapminder %>% 
  filter(year == 2016 & continent == "Europe") %>% 
  slice_max(life_expectancy, n = 5)
```

Bar graph of life expectancy in Europe by countries.

```{r}
gapminder %>% 
  filter(year == 2016 & continent == "Europe") %>% 
  ggplot(aes(y = reorder(country, life_expectancy), x = life_expectancy)) +
  geom_bar(stat = "identity") + 
  coord_cartesian(xlim = c(70,85)) + 
  xlab("Life Expectancy") + ylab(NULL) + 
  labs(title = "Life Expectancy in European Countries (2016)", caption = "Gapminder data.")
```

The GDP of middle-eastern countries.

```{r}
gapminder %>% 
  filter(country %in% c("Israel", "Lebanon", "Egypt", "Saudi Arabia", "Bahrain", "West Bank and Gaza", "Yemen", "United Arab Emirate", "Iran", "Syria")) %>% 
  filter(year == 2009) %>% 
  ggplot(aes(gdp,country)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Bar Graph of GDP in Middle-Eastern Countries") + 
  xlab("GDP") + 
  ylab("Country")
```

What was the correlation between fertility and life expectancy in 1962?

```{r}
gapminder %>% 
  filter(year == 1962) %>% 
  ggplot(aes(fertility, life_expectancy)) + 
  geom_point() + 
  ggtitle("Fertility and Life Expectancy in 1962") + 
  xlab("Fertility") + 
  ylab("Life Expectancy") + 
  theme_clean()
```

What was the correlation between fertility and life expectancy by continents in 1962?

```{r}
gapminder %>% 
  filter(year == 1962) %>% 
  ggplot(aes(fertility, life_expectancy, color = continent)) + 
  geom_point() + 
  ggtitle("Fertility and Life Expectancy in 1962") + 
  xlab("Fertility") + 
  ylab("Life Expectancy")
```

How was it in 2012 compared to 1962?

```{r}
gapminder %>% 
  filter(year %in% c(1962,2012)) %>% 
  ggplot(aes(fertility, life_expectancy, color = continent)) + 
  geom_point() + 
  facet_grid(.~year) + 
  ggtitle("Fertility and Life Expectancy in 1962 and 2012") + 
  xlab("Fertility") + 
  ylab("Life Expectancy")
```

Show me its development in detail over time

```{r}
gapminder %>% 
  filter(year %in% c(1962,1972,1982,1992,2002,2012)) %>% 
  ggplot(aes(fertility, life_expectancy, color = continent)) + 
  geom_point() + 
  facet_wrap(.~year) + 
  ggtitle("Fertility and Life Expectancy from 1962 to 2012") + 
  xlab("Fertility") + 
  ylab("Life Expectancy")
```

What is the fertility distribution in Europe like?

```{r}
gapminder %>% 
  filter(continent == "Europe" & year == 2015) %>% 
  ggplot(aes(fertility, fill = region)) + 
  geom_boxplot() + 
  coord_flip() +
  ggtitle("Fertility Rates in Europe by Regions in 2015") + 
  xlab("Fertility Rates")
```

What is the fertility distribution in Asia like?

```{r}
gapminder %>% 
  filter(continent == "Asia" & year == 2015) %>% 
  ggplot(aes(fertility, fill = region)) + 
  geom_boxplot() + 
  coord_flip() +
  ggtitle("Fertility Rates in Asia by Regions in 2015") + 
  xlab("Fertility Rates")
```

What is the fertility distribution like by continents?

```{r}
gapminder %>% 
  filter(year == 2015) %>%  
  ggplot(aes(fertility,continent, fill = continent)) + ggridges::geom_density_ridges(show.legend = F) + 
  ggtitle("Fertility Rates by Continents in 2015") + 
  xlab("Fertility Rates") + 
  ylab("Continents")
```

Density rigdes on other variable distributions

```{r}
gapminder %>% 
  filter(year == 2015) %>% 
  ggplot(aes(life_expectancy,continent, fill = continent)) + 
  ggridges::geom_density_ridges(show.legend = F) + 
  ggtitle("Life Expectancy by Continents in 2015") + 
  xlab("Life Expectancy") + 
  ylab("Continents")
```

How has the life expectancy in countries by continents changed between the years 1962 and 2012?

```{r}
gapminder %>% 
  filter(year %in% c(1962,2012)) %>% 
  mutate(year = factor(year, levels = c(1962,2012))) %>% 
  ggplot(aes(continent,life_expectancy, fill = year)) + 
  geom_boxplot() + 
  xlab("Continent") + ylab("Life Expectancy") + 
  labs(title = "Life expectancy between continents in 1962 and 2012")
```

## European Social Survey Dataset

```{r cache=TRUE}
library(essurvey)

# Set email. Needed for authentication
set_email("dino.c@me.com")

# # Show which countries are available
show_countries()

# Download data
ess_9 <- import_rounds(9)
```

Skim to get an overview of the variables.

```{r}
# skimr::skim(ess_9)
```

Glimpse to get a peak at the first values.

```{r}
head(ess_9)
```

**Interesting variables**

-   [wltdffr: Differences in wealth in country, how fair](http://nesstar.ess.nsd.uib.no/webview/index.jsp?v=2&submode=variable&study=http%3A%2F%2F129.177.90.83%3A-1%2Fobj%2FfStudy%2FESS9e03.1&gs=undefined&variable=http%3A%2F%2F129.177.90.83%3A80%2Fobj%2FfVariable%2FESS9e03.1_V518&mode=documentation&top=yes)

-   [poltran: Decisions in country politics are transparent](http://nesstar.ess.nsd.uib.no/webview/index.jsp?v=2&submode=variable&study=http%3A%2F%2F129.177.90.83%3A-1%2Fobj%2FfStudy%2FESS9e03.1&gs=undefined&variable=http%3A%2F%2F129.177.90.83%3A80%2Fobj%2FfVariable%2FESS9e03.1_V518&mode=documentation&top=yes)

-   [grspfr: Would you say your gross pay is unfairly low, fair, or unfairly high](http://nesstar.ess.nsd.uib.no/webview/index.jsp?v=2&submode=variable&study=http%3A%2F%2F129.177.90.83%3A-1%2Fobj%2FfStudy%2FESS9e03.1&gs=undefined&variable=http%3A%2F%2F129.177.90.83%3A80%2Fobj%2FfVariable%2FESS9e03.1_V518&mode=documentation&top=yes)

-   [imprich: Important to be rich, have money and expensive things](http://nesstar.ess.nsd.uib.no/webview/index.jsp?v=2&submode=variable&study=http%3A%2F%2F129.177.90.83%3A-1%2Fobj%2FfStudy%2FESS9e03.1&gs=undefined&variable=http%3A%2F%2F129.177.90.83%3A80%2Fobj%2FfVariable%2FESS9e03.1_V518&mode=documentation&top=yes)

-   [ipeqopt: Important that people are treated equally and have equal opportunities](http://nesstar.ess.nsd.uib.no/webview/index.jsp?v=2&submode=variable&study=http%3A%2F%2F129.177.90.83%3A-1%2Fobj%2FfStudy%2FESS9e03.1&gs=undefined&variable=http%3A%2F%2F129.177.90.83%3A80%2Fobj%2FfVariable%2FESS9e03.1_V518&mode=documentation&top=yes)

-   [ipstrgv: Important that government is strong and ensures safety](http://nesstar.ess.nsd.uib.no/webview/index.jsp?v=2&submode=variable&study=http%3A%2F%2F129.177.90.83%3A-1%2Fobj%2FfStudy%2FESS9e03.1&gs=undefined&variable=http%3A%2F%2F129.177.90.83%3A80%2Fobj%2FfVariable%2FESS9e03.1_V518&mode=documentation&top=yes)

How many subjects were questioned per country?

```{r}
ess_9 %>% 
  group_by(cntry) %>% 
  count(sort = T) %>% 
  ggplot() + 
  aes(n, reorder(cntry, n), label = n) + 
  geom_bar(stat = "identity", fill = "steelblue") +
  ggtitle("Subjects per Country")
```

What are the opinions on differences in wealth?

```{r}
ess_9 %>% 
  count(wltdffr) %>% 
  ggplot() + 
  aes(n, wltdffr) + 
  geom_bar(stat = "identity") + 
  ggtitle("Differences in wealth in country, how fair") + 
  ylab("Answers")
```

Here are the results for Germany.

```{r}
ess_9 %>% 
  filter(cntry == "DE") %>% 
  count(wltdffr) %>% 
  ggplot() + 
  aes(n, wltdffr) + 
  geom_bar(stat = "identity") + 
  ggtitle("Differences in wealth in Germany, how fair") + 
  ylab("Answers")
```

What are the options and how many are there?

```{r}
ess_9 %>% 
  distinct(wltdffr)

ess_9 %>% 
  count(wltdffr)
```

Recode values to only have four left at the end, too large unfair or fair and too small unfair and NAs.

```{r}
ess_9 = ess_9 %>% 
  mutate(new_wltdffr = fct_collapse(wltdffr, 
             Too_Large = c("Large, extremely unfair", "Large, very unfair", "Large, somewhat unfair","Large, slightly unfair"), 
             Fair = "Fair", 
             Too_Small = c("Small, extremely unfair", "Small, very unfair", "Small, somewhat unfair","Small, slightly unfair"))) %>% 
  mutate(new_wltdffr = factor(new_wltdffr, levels = c("Too_Large", "Fair", "Too_Small")))
```

Create a count and percentage table.

```{r}
d = ess_9 %>% 
  group_by(new_wltdffr) %>% 
  summarise(count = n()) %>% 
  mutate(perc = count/sum(count))
```

Plot a bar graph.

```{r}
d %>% 
  ggplot(aes(new_wltdffr, perc, label = round(perc, 2), fill = new_wltdffr)) + 
  geom_bar(stat = "identity") + 
  geom_label(aes(fill = NA), fill = "white") + 
  theme(legend.position = "none") + 
  xlab("Differences in Wealth, Opinon") + ylab("Percentage") + 
  ggtitle("Opinion on Differences in Wealth 2018")
```

## WDI Dataset

```{r}
library(WDI)
```

Search for topics in datasets.

```{r cache=TRUE}
head(WDIsearch("co2"), 10)

WDIsearch('gdp.*capita.*constant')

# Download it
dat = WDI(
  country = "all", 
  indicator = c(
  population =  "SP.POP.TOTL", 
  gdp = "NY.GDP.MKTP.CD",
  inc_sha_10 = "SI.DST.10TH.10"),
  start = 1960, end = 2018)
```

What was the population development like in these specific countries?

```{r}
dat %>% 
  filter(country %in% c("Germany","France","United States", "Bosnia and Herzegovina","United Kingdom","China")) %>% 
  ggplot(aes(year,population, color = country)) + 
  geom_line() + 
  scale_y_log10(labels = scales::comma) + 
  ylab("Population (log10)") + xlab("Year") + 
  labs(title = "Random Countries Population", color = "Country")
```

What was the population development like in the former Yugoslav Republics?

```{r}
dat %>% 
  filter(country %in% c("Bosnia and Herzegovina","Croatia","Serbia","Montenegro","Slovenia","North Macedonia","Kosovo")) %>% 
  ggplot(aes(year,population, color = reorder(country, desc(population)))) + 
  geom_line() + 
  scale_x_continuous(breaks = seq(1960,2020,5)) + 
  scale_y_continuous(labels = scales::comma) + 
  labs(title = "Population of Former Yugoslav Republics", subtitle = "Population trends and shifts since the 1990s", col = "Countries") + 
  xlab("Year") + ylab("Population")
```

What was the development of GDP in the former Yugoslav Republics?

```{r}
dat %>% 
  filter(country %in% c("Bosnia and Herzegovina","Croatia","Serbia","Montenegro","Slovenia","North Macedonia","Kosovo")) %>% 
  filter(between(year, 1995, 2020)) %>% 
  ggplot(aes(year,gdp, color = reorder(country, desc(gdp)))) + 
  geom_line() + 
  scale_x_continuous(breaks = seq(1960,2020,2)) + 
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "GDP of Former Yugoslav Republics", subtitle = "Economic trends and shifts since the 1990s" , col = "Countries") +
  xlab("Year") + ylab("GDP current US$")
```

# 2. Frequent Data Analysis Problems + Solutions

## First exploration of new dataset

The `skim()`shows how many missing and unique values each variable has. It uses appropriate measures to describe each variable based on its type: character, numeric or list.

```{r}
skimr::skim(starwars)
```

The glimpse function, on the other hand, gives us a good peak at the first raw values each variable has.

```{r}
glimpse(starwars)
```

## A Count and prop table

**First way with `forcats::fct_count()`** Calculates a count and prop table.

```{r}
starwars$sex %>%
 factor() %>% 
  fct_count(sort = T, prop = T)
```

**Second way with `deplyr::count()`** Simply mutate a frequency and percentage column on a counted table.

```{r}
starwars %>% 
 count(sex) %>% 
  mutate(freq = n / sum(n)) %>% 
  mutate(perc = freq * 100)
```

## Bar graph with count data

Here is a situation where we calculated a count table for hair color - we summarized all values. If we then want to plot a bar graph based on that count table we run into problems, because ggplot2 is expecting a non-summarized or normal data frame.

```{r}
hair_color_table = starwars %>% 
  mutate(hair_color = fct_lump_min(hair_color, 2)) %>% 
  group_by(hair_color) %>% 
  summarise(n = n())
hair_color_table
```

To tell the function that we have already summarized data, we add the argument `stat = "identity"` to the `geom_bar()` function.

```{r}
hair_color_table %>% 
  ggplot(aes(x = reorder(hair_color, n), y = n, fill = hair_color)) + 
  geom_bar(stat = "identity") + 
  theme(legend.position = "none")
```

## Bar graph with percentage labels

First we create a table with counts and percentages:

```{r}
d = starwars %>% 
  group_by(gender) %>% 
  summarise(count = n()) %>% 
  mutate(percentage = count/sum(count))
d
```

Then we plot a graph with bar and with percentage labels.

```{r}
d %>% 
  ggplot(aes(gender, percentage, label = round(percentage, 2), fill = gender)) + 
  geom_bar(stat = "identity") + 
  geom_label(aes(fill = NA), fill = "white") + 
  theme(legend.position = "none")
```

## Collapse factors to „Other"

This syntax mutates the categorical variable homeworld into eight of its most frequent values. The other values are being collapsed into the categorical value „other".

```{r}
starwars %>% 
  mutate(homeworld = fct_lump_n(homeworld, n = 8)) %>% 
  group_by(homeworld) %>% 
  summarise(mean(height, na.rm =T), mean(mass, na.rm = T), n())
```

## Filter for specific values

We can easily filter out cases with certain column values, like for example the states of Hawai and Alaska. We use `filter()`, the operator `!` and `%in%`.

```{r}
starwars %>% 
  filter(!homeworld%in%c("Tatooine","Naboo")) %>% 
  select(name, homeworld)
```

## Change bar colors in barplot

You can manually pick the colors with `fill` and a vector containing the color values. Either in String, written out.

```{r}
starwars %>% 
  mutate(sex = fct_infreq(sex)) %>% 
  ggplot(aes(sex)) + 
  geom_bar(fill = c("red","blue","green","black","grey")) 
```

Or with RGB Color Codes.

```{r}
starwars %>% 
    mutate(sex = fct_infreq(sex)) %>%
    ggplot(aes(sex)) +
    geom_bar(fill = c("#003f5c","#58508d","#bc5090","#ff6361","#ffa600")) 
```

## Hide aes(color) mapping legend

Here is an example where we want the bar colored based on the variable itself, but without the mapping legend.

```{r}
starwars %>% 
  mutate(sex = fct_infreq(sex)) %>% 
  ggplot(aes(sex, fill = sex)) + 
  geom_bar()
```

Hide the geom_bar legend.

```{r}
starwars %>% 
  mutate(sex = fct_infreq(sex)) %>% 
  ggplot(aes(sex, fill = sex)) + 
  geom_bar(show.legend = F)
```

Remove just the legend title:

```{r}
starwars %>% 
  mutate(sex = fct_infreq(sex)) %>% 
  ggplot(aes(sex, fill = sex)) + 
  geom_bar() +
  theme(legend.title = element_blank())
```

Hide all legends created:

```{r}
starwars %>% 
  mutate(sex = fct_infreq(sex)) %>% 
  ggplot(aes(sex, fill = sex)) + 
  geom_bar() +
  theme(legend.position = "none")
```

## Re-code values of categorical variables

**First way** We can use `fct_collapse()`to create a new column with the new recoded values in it.

```{r eval=FALSE, include=FALSE}
ess_9 = ess_9 %>% 
  mutate(new_wltdffr = fct_collapse(wltdffr, 
             Too_Large = c("Large, extremely unfair", "Large, very unfair", "Large, somewhat unfair","Large, slightly unfair"), 
             Fair = "Fair", 
             Too_Small = c("Small, extremely unfair", "Small, very unfair", "Small, somewhat unfair","Small, slightly unfair"))) %>% 
  mutate(new_wltdffr = factor(new_wltdffr, levels = c("Too_Large", "Fair", "Too_Small")))
```

**Second way** By using `mutate`, to create a new column with our own values and `case_when`, to run through our observations looking for defined cases, together with "variable" `%in%`, we can create our own groups.

```{r}
gapminder %>% 
 mutate(group = case_when(
    region %in% c("Western Europe", "Northern Europe","Southern Europe","Northern America", "Australia and New Zealand") ~ "West", # If region is one of values -> assign it "West" in new group column.
    region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
    region %in% c("Caribbean", "Central America", "South America") ~ "Latin America",
    continent == "Africa" & 
      region != "Northern Africa" ~ "Sub-Saharan",
    TRUE ~ "Others")) %>%  # If nothing above applies -> assign it "Others" in group column
  head(10)

```

We turn this `group` variable into a factor to control the order of the levels:

```{r eval=FALSE, include=FALSE}
gapminder %>% 
  mutate(group = factor(group, levels = c("Others", "Latin America", "East Asia", "Sub-Saharan","West")))
```

## Order color legend

Order color legend by a variable's values.

```{r eval=FALSE, include=FALSE}
data %>% 
    ggplot(aes(year,population, color = reorder(country, desc(population)))) + # "Reorder countries by descending order of population"
    geom_line() 
```

## Show unique values

Display all unique values of variable.

```{r}
distinct(starwars, species) # dplyr function
```

**Note**: `distinct(dat$countries)` doesn't work.

## Slice rows by maximum or minimum values

**Note:** parameter `n` must be explicitly written, otherwise it throws an error.

```{r}
starwars %>% 
  slice_max(height, n = 5)
```

Show me 5% of the lowest height rows.

```{r}
starwars %>% 
  slice_min(height, prop = 0.05)
```

## Show Number of `NA`s

For a quick check of how many missing values there are in a single column:

```{r}
sum(is.na(starwars$height))
```

And how many are not `NA`s.

```{r}
sum(!is.na(starwars$height))
```

For a more detailed overview of the whole dataset use `skim()`. It shows a very useful `complete_rate` which tells us how much of the column is disturbed by missing values.

```{r}
skimr::skim(starwars)
```

## Drop rows with missing values

Drop rows that have `NA`values in a specific column, here in `height`.

```{r}
starwars %>% 
  drop_na(height)
```

Drop all rows that contain `NA` in any column.

```{r}
starwars %>% 
  drop_na()
```

Filter out any NA containing rows.

```{r}
starwars %>% 
 na.exclude()
```

## Replace `NA`s

Replace 0 with value you want as a replacement.

```{r}
data(na_example)
sum(is.na(na_example))

no_nas <- ifelse(is.na(na_example), 0, na_example) # "if is NA is true, change value to 0, else keep the value (i.e. na_example)"

sum(is.na(no_nas))
```

## The factor variable trap

The FVT is about what happens when you try to return factorized vectors into numeric values. Let's look at this with this code.

```{r}
z <-factor(c("12", "13", "14", "15", "12")) # We create an object by directly factorizing a vector. 
z

y <- as.numeric(z) # Now we want to convert them into numeric values. 
y # What?
```

This happened, because we picked up the on the factorization result. `factor()` assigns every element, based on its value, an integer number.

```{r}
typeof(z) # 1=12, 13=2, 14=3, 15=4, 12=1
```

To fix this problem, first convert the object back to character and then to numeric.

```{r}
y <- as.numeric(as.character(z))
y
```
